A Systematic Literature Review on Extensions to the Role-Based Access Control Reference Model

### 2014-06-17 paper/sections/analysis.tex
Revised version
 - Address comments in background, RQ2-RQ5, and discussion section.
 - In addition, I sort paper list in an increasing order (e.g., [4][5][10] instead of [10][4][5]) in the table.

\section{RQ2: Motivations} \label{sec:motivations}

\textit{What are motivations behind the RBAC extension models?}
\\

During the paper reading phase, we identify motivations of the RBAC extension models by category. 
We identify motivations by looking for issues regarding why the RBAC reference model was inadequate and how the authors addressed the issues. 
For example, the papers in the Spatial Category argue that the RBAC reference model does not support spatial constraints, which are used in determining users' access decisions based on spatial dimensions (i.e., location). 
To address this issue, the authors developed the RBAC extension models by incorporating additional spatial entities and their relations with existing entities of the RBAC Reference Model.

We describe motivations behind the RBAC extension models by category as follows:

\begin{itemize}

  \item \textbf{Context}: The RBAC Reference Model does not support the notion of context constraints related to changes in environments. Therefore, RBAC belongs to the static access control model, which may not capture changes in environments.

  \item \textbf{Constraint}: The RBAC Reference Model has limitations on its features such as delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.

  \item \textbf{Organizational}: The RBAC extension model does not handle RBAC administrative tasks efficiently across multiple organizations. The model needs to reduce the administrative complexity of RBAC across multiple organizations.

  \item \textbf{Privacy}: The RBAC extension model does not support the notion of privacy. For example, the model lacks components, constraints, and obligations to handle privacy in RBAC.

  \item \textbf{Task}: The RBAC extension model does not support the notion of a task, team, purpose, and organizational roles, which help specify a business activity in enterprises.

  \item \textbf{Spatio-Temporal}: The RBAC extension model does not support spatial (location-based) and temporal (time-based) constraints, which specify role-assignment, role-activation and permissions based on location and time.

  \item \textbf{Spatial}: The RBAC extension model does not support the notion of spatial (location-based)  constraints, which specify role-assignment and permissions based on location.

  \item \textbf{Temporal}: The RBAC extension model does not support the notion of temporal (time-based)  constraints, which specify role-assignments and permissions based on time.

\end{itemize}

The RBAC standard provides the RBAC Reference Model, which is an abstraction on top of authorization based on $UA$ and $PA$. RBAC Reference Model presents four entities: roles, sessions, users, and permissions, and their relations. While the RBAC Reference Model is considered fundamental in any RBAC systems, the RBAC Reference Model has limitations on providing features such as dealing with context for emerging applications such as healthcare and mobile devices. For example, in cases where the RBAC extension model does not support the notion of specific constraints, the RBAC extension model incorporates additional entities with regards to context/constraints and their relations with existing entities of the RBAC Reference Model. The RBAC extension model provides general agreement on the definition for specific features. The RBAC extension model helps simplify theoretical modeling and practical implementation of features in the model.

%The Const category claims that the RBAC extension model has limitations on its features. The RBAC extension model in this category incorporates modification of existing features in the RBAC Reference Model. For models in this category, changes occur within the RBAC Reference Model.
%The Task category claims that the RBAC extension model does not support the notion of task, team, purpose, and organizational roles. Because task-based access control has been developed, the RBAC extension model in this category is a combination of the RBAC Reference Model and another access control model, which uses model-specific attributes and their relations. For this combination, new associations between attributes of two models should be introduced. Models in this category require new associations between attributes of different access control models to work together.
%
%
%The Context, Org, Priv, Spatio-Temporal, Spatial, and Temp categories claim that the RBAC extension model does not support the notion of specific constraints. The RBAC extension model in these categories incorporates additional entities with regards to context/constraints and their relations with existing entities of the RBAC Reference Model. The extension model does not change the RBAC Reference Model.
%
%The Const category claims that the RBAC extension model has limitations on its features. The RBAC extension model in this category incorporates modification of existing features in the RBAC Reference Model. For models in this category, changes occur within the RBAC Reference Model.
%
%The Task category claims that the RBAC extension model does not support the notion of task, team, purpose, and organizational roles. Because task-based access control has been developed, the RBAC extension model in this category is a combination of the RBAC Reference Model and another access control model, which uses model-specific attributes and their relations. For this combination, new associations between attributes of two models should be introduced. Models in this category require new associations between attributes of different access control models to work together.

\greybox{Motivations for RBAC extension models vary across category but are rooted in adding new entities or relationships to allow authorization flexibility in meeting the demands of emerging requirements.}

\section{RQ3: Implementations} \label{sec:implementations}

@@

When designing and proposing a model targeted at a feature that is rooted in practical
usage by real software systems, implementing the model is evidence that the
proposed model can work in practice. The presence of an actual implementation substantiates the robustness of the design and/or the need for the extension.
We analyzed the primary sources to see how many
proposed models actually had implementations associated with them.
We quantified whether the implementation was enterprise (i.e., an implementation used in a production environment) was for a real system) or a prototype implementation (i.e., proof of concept to demonstrate the feasibility).

@@

\section{RQ4: Evaluations} \label{sec:evaluations}

@@

The 27 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors. We identified a set of evaluation types found within  each of the primary sources and provide a list of the evaluation types and which primary sources provided which type. A primary source may use multiple evaluation types. For example, Aich et al. conducted performance and complexity analyses to show the effectiveness of their proposed RBAC extension model \cite{aich09:role}.

@@

Table \ref{tab:evaluations} shows evaluation types.
Based on the diverse evaluation criteria, 12 models presented no evidence of an evaluation. Fourteen models presented example scenarios
and how application of their model would apply and resolve the situation. Six models provided some form of performance
or complexity analysis of their model.  The performance and complexity analysis included graphs of the model's time to determine calculate authorization 
as the number of entities grew, and the size of the role space for the extension model compared to the RBAC Reference Model. 
Five models provided mathematical descriptions and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation approach was providing sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios. 
%Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
%or show how the Reference Model is deficient in tackling said scenarios.

\begin{table}
\centering
\caption{Evaluation types by primary source}
\vspace{0.1 in}
\begin{tabular*}{.9\linewidth}{| p{.45\linewidth} | p{.3\linewidth} | r | }
\cline{1-3}

\textbf{Evaluation Type} & \textbf{Papers} & \textbf{Count} \\ \cline{1-3}
Time-Based Performance
&
\cite{ni2010privacy}
\cite{aich09:role}
&
2 \\ \cline{1-3}
Complexity Analysis
&
\cite{bao08:role}
\cite{zhang06:collaborative}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
&
4 \\ \cline{1-3}
Comparison to Standard RBAC
& 
\cite{bao08:role}
\cite{zou2009crbac}
\cite{zhang06:collaborative}
\cite{ray07:spatio}
\cite{zhao2008flexible}
&
5 \\ \cline{1-3}
Mathematical Modeling
& 
\cite{damiani2007geo}
\cite{hansen2003spatial}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
\cite{joshi05:generalized}
&
5 \\ \cline{1-3}
Example Scenarios of the Model in Action
&
\cite{jian2008extended}
\cite{yao2008task}
\cite{cholewka00:acontext-sensitive}
\cite{huang06:pervasive}
\cite{bao08:role}
\cite{zhou2007team} 
\cite{alam06:constraint}
\cite{tzelepi01:flexible}
\cite{yamazaki04:designing}
\cite{zou2009crbac}
\cite{samuel07:spatio-temporal}
\cite{ray07:spatio}
\cite{joshi05:generalized}
\cite{oh2003task}
&
14 \\ \cline{1-3}
Case Study of the Model in Practice
&
\cite{motta03:contextual}
&
1 \\ \cline{1-3}
None
&
\cite{masoumzadeh2008purbac}
\cite{ray2006lrbac}
\cite{chandran05:llt}
&
3 \\ \cline{1-3}

@@

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. 
For five of the primary sources \cite{bao08:role} \cite{zou2009crbac} \cite{zhang06:collaborative} \cite{zhao2008flexible} \cite{ray07:spatio} the model authors provide some discussion of how the RBAC Reference Model is deficient in tackling the scenario.
In Table \ref{tab:evaluations}, we observe that one paper includes a case study to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC Reference Model, developers may be left implementing a more complex model to address their requirements when the RBAC Reference Model would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

When looking for an enterprise ready access control mechanism, developers must balance usability with security. 
Two of the primary sources examined provided time-based performance analysis of their extension model compared
to the RBAC Reference Model. This inclusion of time analysis provides some assurances to developers that any non-functional
requirements surrounding time to compute authorizations compete or beat the RBAC Reference Model. Further, four of the models provided
some form of complexity of their model. This complexity analysis plays a key role in the management of the access control mechanism
over the course of the models implementation lifetime. As the number of roles, users and additional entities grows, developers will 
need to ensure non-functional requirements are met that deal with the ability for a system administrator to effectively manage these
entities.

@@

\section{RQ5: Domains} \label{sec:domains}

@@

\section{RQ6: Generalizations} \label{sec:generalizations}

@@

The RBAC standard is used in various aspects of computer systems. To reduce efforts for modeling access control used in various applications, researchers often focus on developing generalized core concepts of access control.
The authors use propositional logic to describe access control model across all categorizations. Propositional logic is concerned with propositions and their logical relationships. In propositional logic, simple (i.e., atomic) or compound condition at given context is evaluated to true or false based on specified rules and access control logic. Researchers and practitioners are concerned to extend limited set of propositions specific to core RBAC to meet real-world scenarios such as dynamic constraints, temporal, or spatial constraints.

@@

Since NIST proposed RBAC standard using propositional logic, researchers describe extended RBAC models using propositional logic. Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true or false based on predicates, propositional logic is sufficient to describe key ideas and definitions. Propositional logic specifies what combination of attributes values a request must satisfy to be authorized to roles. 
Combination of attribute values may handle constraints such as temporal and spatial constraints. For example, we describe user-role authorization as $ae$ $\Rightarrow$ $r$ using propositional logic where $ae$ is an attribute expression and $r$ is a role. Propositional logic for RBAC has two parts. The left hand side (LHS) is an attribute expression $ae$. 
The right hand side (RHS) is a role. 
If a request satisfies the attribute expression, a user $u$ in the request is authorized to the role specified in RHS.

We found that 27 papers of RBAC extensions use propositional logic to describe its extended model.
However, propositional logic has limitations. While this logic is simple, this logic does not support for reasoning about RBAC, which helps
reduce the administrative complexity of associations such as user- role associations.
To support for reasoning of access control, one may describe the RBAC extended models in first-order logic. 
Given an RBAC extension model, Samuel et al.~\cite{samuel07:spatio-temporal} proposed verification of the model using a specification language, which is based first-order logic. 
This logic is sufficient to model RBAC and extended RBAC for reasoning. 
Moreover, this logic supports for concise and elegant formulation of the Reference Model and its relation. 
First-Order logic is expressive enough to concisely represent access control systems. First-Order logic uses relations, variables, and quantifiers.

>>>>

\section{RQ2: Implementations} \label{sec:implementations}

@@

When designing and proposing a model targeted at a feature that is rooted in practical
usage by real software systems, implementing the model is evidence that the
proposed model can work in practice. The presence of an actual implementation substantiates the robustness of the design and/or the need for the extension.
We analyzed the primary sources to see how many
proposed models actually had implementations associated with them.
We quantified whether the implementation was an enterprise implementation (i.e., an implementation used in a production environment for a real system) or a prototype implementation (i.e., proof of concept to demonstrate the feasibility). In scientific research software life cycle, NuNamaker et al.~\cite{nunamaker1990systems} illustrated three implementation stages: prototype, enterprise (i.e., product), and technology transfer. Researchers often conduct their research by building a prototype implementation. A prototype implementation serves as a proof of concept to verify that certain concepts of proposed RBAC extension models. The prototype implementation demonstrates feasibility. However, this prototype implementation does not represent a deliverable real-world application. Majority of implementations in RBAC research stops with prototyping. After prototypes are successfully demonstrated, researchers make an effort to further develop a prototype implementation into an enterprise implementation. An enterprise implementation is a deliverable real-world application and allows to conduct realistic evaluations of the RBAC extension models. Technology transfer means that the technology of the proposed RBAC extension models are eventually transferred to the public marketplace. We found that none of implementations of the primary sources reach to the stage of the transfer of technology.

@@

\section{RQ3: Evaluations} \label{sec:evaluations}

@@

The 27 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors. 
Zelkowitz and Wllace~\cite{Zelkowitz1998experimentalmodels} proposed evaluation types and methods that are used to validate the claims in the paper. 
Table~\ref{tab:evaluationtypes} illustrates 12 evaluation methods with their corresponding general evaluation types and descriptions. We classified each paper according to the evaluation methods shown in Table~\ref{tab:evaluationtypes}. For example, Aich et al. conducted project monitoring to show the performance and complexity of their proposed RBAC extension model \cite{aich09:role}.

@@

Table~\ref{tab:evaluations} shows evaluation methods and criteria.
Based on the diverse evaluation methods, 12 models presented no evidence of an evaluation. Fourteen models presented 
example scenarios and how application of their model would apply and resolve the situation. These methods refer to assertions.
Six models provided some form of performance
or complexity analysis of their model through project monitoring.  The performance and complexity analysis included graphs of the model's time to determine calculate authorization 
as the number of entities grew, and the size of the role space for the extension model compared to the RBAC Reference Model. 
Five models provided simulations based on mathematical descriptions and analysis as a way to provide evaluation in the form of completeness. 
The most widely used evaluation method was assertion, which provides sample scenarios with accompanying workflows of how the extension model
would tackle those scenarios.
%Much is left to the reader to assume of these types of evaluations, as the authors do not explicitly state
%or show how the Reference Model is deficient in tackling said scenarios.

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[t]
  \centering
  \caption{Evaluation Types and methods classified by Zelkowitz and Wllace~\cite{Zelkowitz1998experimentalmodels}}
  %  \begin{tabular}{|l|l|l|}
   \begin{tabular*}{.9\linewidth}{| p{.25\linewidth}| p{.2\linewidth} | p{.35\linewidth} |}
		\hline

    \textbf{Evaluation Type} & \textbf{Method} & \textbf{Description} \\\hline

    \multirow{4}[8]{*}{Observational methods} & Project monitoring & Monitor project and collect development data \\\cline{2-3}
          & Case study & Monitor project in depth and collect data with a specific goal for the project \\\cline{2-3}
          & Assertion & This evaluation uses examples for validation\\\cline{2-3}
          & Field study & Monitor multiple projects and collect relevant information to compare the projects \\\hline
    \multirow{3}[6]{*}{Historical methods} & Literature research & Examine previously published projects \\\cline{2-3}
          & Legacy & Examine data from completed projects \\\cline{2-3}
          & Lessons learned & Examine qualitative data from projects \\\hline
    \multirow{5}[10]{*}{Controlled methods} & Static analysis & Examine structure of dveloped product \\\cline{2-3}
          & Replicated experiment & Develop mulitple versions of product \\\cline{2-3}
          & Synthetic environment & Replicate one factor in laboratory setting \\\cline{2-3}
          & Dynamic analysis & Execute developed product for perormance \\\cline{2-3}
          & Simulation & Execute product using a model of real enviroment \\\hline

    \end{tabular*}%
  \label{tab:evaluationtypes}%
\end{table}%

\begin{table}
\centering
\caption{Evaluation methods by primary source}
\vspace{0.1 in}
\begin{tabular*}{.9\linewidth}{| p{.2\linewidth}| p{.35\linewidth} | p{.2\linewidth} | r | }
\cline{1-4}
\textbf{Methods} & \textbf{Criteria} & \textbf{Papers} & \textbf{Count} \\ \cline{1-4}
\multirow{2}[0]{*}{Project Monitoring}
&
Time-Based Performance
&
\cite{ni2010privacy}
\cite{aich09:role}
&
2 \\ \cline{2-4}

&
Complexity Analysis
&
\cite{bao08:role}
\cite{zhang06:collaborative}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
&
4 \\ 
\cline{1-4}
Case Study
&
RBAC Model in Practice
&
\cite{motta03:contextual}
&
1 \\ \cline{1-4}
Assertion
&
Example Scenarios of the Model in Action
&
\cite{jian2008extended}
\cite{yao2008task}
\cite{cholewka00:acontext-sensitive}
\cite{huang06:pervasive}
\cite{bao08:role}
\cite{zhou2007team} 
\cite{alam06:constraint}
\cite{tzelepi01:flexible}
\cite{yamazaki04:designing}
\cite{zou2009crbac}
\cite{samuel07:spatio-temporal}
\cite{ray07:spatio}
\cite{joshi05:generalized}
\cite{oh2003task}
&
14 \\ \cline{1-4}
Field Study
&
Comparison to Standard RBAC
&  
\cite{bao08:role}
\cite{zou2009crbac}
\cite{zhang06:collaborative}
\cite{ray07:spatio}
\cite{zhao2008flexible}
&
5 \\ \cline{1-4}
Simulation
&
Mathematical Modeling
&  
\cite{damiani2007geo}
\cite{hansen2003spatial}
\cite{aich07:STARBAC}
\cite{chen08:spatio-temporal}
\cite{joshi05:generalized}
&
5 \\ \cline{1-4}
No Evaluations
&
Not Applicable
&
\cite{masoumzadeh2008purbac}
\cite{ray2006lrbac}
\cite{chandran05:llt}
&
3 \\ \cline{1-4}

@@

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. 
For five of the primary sources \cite{bao08:role} \cite{zou2009crbac} \cite{zhang06:collaborative} \cite{ray07:spatio} \cite{zhao2008flexible}, the model authors conduct field study to provide some discussion of how the RBAC Reference Model is deficient in tackling the scenario.
In Table \ref{tab:evaluations}, we observe that one paper includes a case study to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC Reference Model, developers may be left implementing a more complex model to address their requirements when the RBAC Reference Model would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

When looking for an enterprise ready access control mechanism, developers must balance usability with security. 
Two of the primary sources examined provided time-based performance analysis of their extension model compared
to the RBAC Reference Model through project monitoring. This inclusion of time analysis provides some assurances to developers that any non-functional
requirements surrounding time to compute authorizations compete or beat the RBAC Reference Model. Further, four of the models provided
some form of complexity of their model through project monitoring. This complexity analysis plays a key role in the management of the access control mechanism
over the course of the models implementation lifetime. As the number of roles, users and additional entities grows, developers will 
need to ensure non-functional requirements are met that deal with the ability for a system administrator to effectively manage these
entities.

@@

\section{RQ4: Domains} \label{sec:domains}

@@

\section{RQ5: Generalizations} \label{sec:generalizations}

@@

The RBAC standard is used in various aspects of computer systems. To reduce efforts for modeling access control used in various applications, researchers often focus on developing generalized core concepts of access control.
The authors use formal model to provide high level description of access control model across all categorizations.
Formal model of access control provides precise understanding, specification, and analsysi of access control.
Formal model~\cite{abadi2008variations} of access control can be specified in one of logic expressions such as propositional logic. For example, in propositional logic, simple (i.e., atomic) or compound condition at given context is evaluated to true or false based on specified rules and access control logic. Aich et al.~\cite{aich09:role} proposed STARBAC model,  which can be expressed in propositional logic. While propositional logic does not support for quantifiers, first-order logic (i.e., predicate logic) extends propositional logic by the use of quantifiers. For example, ESTARBAC model~\cite{aich09:role} can be expressed in first-order logic.

@@

We found that 27 papers of RBAC extensions provide formal model based on logic expressions to describe its extended model.
%Since NIST proposed RBAC standard using logic expression, we observed that researchers describe extended RBAC models using logic expressions.
We discuss two widely used logic expressions for access control: propositional logic and first-order logic. 
Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true or false based on predicates, propositional logic is sufficient to describe key ideas and definitions. Propositional logic is suitable for specifying what combination of attributes values a request must satisfy to be authorized to roles.
However, propositional logic has limitations. While this logic is simple, this logic does not support for quantifiers and reasoning about RBAC, which helps
reduce the administrative complexity of associations such as user- role associations.
To support for reasoning of access control, one may describe the RBAC extended models in first-order logic. First-order logic is expressive enough to concisely represent access control systems.
Given an RBAC extension model, Samuel et al.~\cite{samuel07:spatio-temporal} proposed verification of the model using a specification language, which is based first-order logic. 
This logic is sufficient to model RBAC and extended RBAC for reasoning. 
Moreover, this logic supports for concise and elegant formulation of the Reference Model and its relation.

%First-order logic uses relations, variables, and quantifiers.

%Combination of attribute values may handle constraints such as temporal and spatial constraints. For example, we describe user-role authorization as $ae$ $\Rightarrow$ $r$ using propositional logic where $ae$ is an attribute expression and $r$ is a role. Propositional logic for RBAC has two parts. The left hand side (LHS) is an attribute expression $ae$. 
%The right hand side (RHS) is a role. 
%If a request satisfies the attribute expression, a user $u$ in the request is authorized to the role specified in RHS.

.

### 2014-06-17 paper/sections/core_rbac.tex
Revised version
 - Address comments in background, RQ2-RQ5, and discussion section.
 - In addition, I sort paper list in an increasing order (e.g., [4][5][10] instead of [10][4][5]) in the table.

RBAC provides effective and efficient permissions management for operations, especially when sharing resources within an organization.
Prior to the creation of the NIST RBAC standard, no general agreement on the definition 
of RBAC existed among practitioners or within the research community. 
Without a unified definition of RBAC, software developers described similar concepts and features of RBAC models using different terminology. 
The lack of consistent terminology was shown to slow the implementation of RBAC~\cite{o20102010}.  
Moreover, in cases where organizations were concerned with adopting RBAC,
evaluation and comparison of RBAC technologies developed by different vendors was difficult.
NIST, in collaboration with industry and academics, worked on defining a set of consensus RBAC concepts and terminology and proposed a standard for 
RBAC that addressed these cost and interoperability issues by developing a common definition that can be used across different vendors.

NIST's work can directly benefit organizations by lowering the cost of early phase research and development, and implementation of RBAC.
Since the RBAC standard was first introduced, a 2010 report by RTI International showed that the rate of RBAC adoption has rapidly grown~\cite{o20102010}. 
The analysts estimate that the use of RBAC technology has saved industry \$6.1 billion from 1992 to 2009 and the research and development work by NIST to produce an RBAC standard has saved an additional \$1.1 billion that wold have otherwise been spent across the industry doing research and development on RBAC.

The RBAC standard includes three components of RBAC: core RBAC, hierarchical RBAC, and constrained RBAC. Each component includes a corresponding RBAC Reference model. We first describe the core RBAC, associated entities and other terminology. We next describe hierarchical RBAC and constrained RBAC, which are developed by incorporating new features into core RBAC.

\subsection{Core RBAC}

The four entities of the core RBAC Reference Model are:

@@

@@

\subsection{Hierarchical RBAC}

The hierarchical RBAC Reference Model adds role hierarchies ($RH$) as a feature to the core RBAC Reference Model.
$RH$ incorporates a structure of roles in an organization using inheritance relationships among attributes such as roles.
The role structure in an organization may use
a role $r_1$, which inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of an employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments for a user. 
Concept of the role inheritance describes the many-to-many mapping role inheritance relations among roles.
Therefore, more than one role (e.g., two roles $r_1$ and $r_1'$) can inherit all permissions of $r_2$ 
General role hierarchies can be extended to use the concept of multiple inheritances where
$r_1$ inherits all permissions from more than one role (e.g., two roles $r_2$ and $r_2'$) .

\subsection{Constrained RBAC}

The constrained RBAC Reference Model adds separation of duty relations to the core and hierarchical RBAC Reference Models. 
Separation of duty relations enforce conflicts of interest among roles. 
The Constrained RBAC model defines two types of separation of duty relations: static and dynamic.

\begin{itemize}
\item Static Separation of Duty (SSoD): SSoD restricts the conflicting-roles, which can be assigned to a single user statically. 
Suppose roles $Role_A$ and $Role_B$ conflict with each other. On situations where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ statically. 
SSoD is known to be too rigid for practical use in cases where a user should have permissions when a user is assigned to both $Role_A$ and $Role_B$. 
For example, an accounting clerk role requests a check and an accounting manager role approves the check. 
These two roles must be mutually exclusive to avoid a situation where one approves the check that the other requested.

\item Dynamic Separation of Duty (DSD): DSD is known to be more flexible than SSoD. 
DSD restricts the conflicting-role assignments dynamically that are associated with a user. 
Suppose roles $Role_A$ and $Role_B$ conflict with each other. 
For situations where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ dynamically.	
Consider that a user can be assigned to the accounting clerk and accounting manager role at the same time.
In such a situation, DSD can enforce that the accounting manager role does not approve her/his requested checks but can only approve checks that others have requested.
\end{itemize}

>>>>

RBAC provides a high level abstraction of permissions management for operations, especially when sharing resources among roles within an organization. 
In cases where organizations were concerned with adopting RBAC, evaluation and comparison of RBAC technologies developed by different vendors was difficult.
To address this issue, NIST proposed a standard for RBAC.
NIST's RBAC standard can benefit organizations by lowering the cost of RBAC adoption~\cite{o20102010}.
The RBAC standard includes three components of RBAC: core RBAC, hierarchical RBAC, and constrained RBAC. 
Hierarchical RBAC and constrained RBAC are developed by incorporating new features into core RBAC.
Each component includes a corresponding RBAC Reference model.

We describe the four entities of these reference models.

@@

%\subsection{Core RBAC} 
%The four entities of the core RBAC Reference Model are:

@@

On the top of the core RBAC Reference Model, the hierarchical RBAC Reference Model adds role hierarchies ($RH$) as a feature. 
The role structure in an organization may use
a role $r_1$, which inherits all permissions of another role $r_2$.
Concept of the role inheritance describes the many-to-many mapping role inheritance relations among roles. Therefore, more than one role (e.g., two roles $r_1$ and $r_1'$) can inherit all permissions of $r_2$.

The constrained RBAC Reference Model adds separation of duty relations to the core and hierarchical RBAC Reference Models. Separation of duty relations enforce conflicts of interest among roles. 
The Constrained RBAC model defines two types of constraints placed on the user-role assignments:
Static Separation of Duty (SSoD) and Dynamic Separation of Duty (DSD).

\begin{itemize}
\item SSoD relations define constraints as a pair ($role set$, $n$) statically where no user is assigned to more than $n$ roles from the $role set$.
Suppose roles $Role_A$ and $Role_B$ in $role set$ conflict with each other and $n$ is 2. A user who is assinged to $Role_A$ cannot be assigned to $Role_B$.
%For example, an accounting clerk role requests a check and an accounting manager role approves the check. 
%These two roles must be mutually exclusive to avoid a situation where one approves the check that the other requested.

\item DSD relations define constraints as a pair ($role set$, $n$) dynamically where $n$ is a number, with the property that
user session may not activate more than $n$ roles from the $role set$.
For situations where multiple roles can be associated with a single user and $n$ is 2, a user session cannot be assigned to both $Role_A$ and $Role_B$ at the same time. 
%Consider that a user can be assigned to the accounting clerk and accounting manager role at the same time.
%In such a situation, DSD can enforce that the accounting manager role does not approve her/his requested checks but can only approve checks that others have requested.
\end{itemize}

### 2014-06-17 paper/sections/discussion.tex
Revised version
 - Address comments in background, RQ2-RQ5, and discussion section.
 - In addition, I sort paper list in an increasing order (e.g., [4][5][10] instead of [10][4][5]) in the table.

The research questions we identified, presented results for and analyzed provide a view of the RBAC extension model landscape. 
By looking across all the research questions we can arrive at cross-cutting concerns and identify areas that may benefit from future work.
We break the discussion into two issues: recommendations for the enhancement of the evaluation of RBAC extension models to role-based access control models and guidelines for developers needing to choose an RBAC extension model to meet their requirements.

\subsection{Metrics for the Evaluation of RBAC Extension Models}

A small percentage of primary sources presented robust evaluations or implementations of their model. 
This result leaves researchers and practitioners little data from which to compare one model to the next. 
Given the wide swath of popular and critical domains being targeted by RBAC extension models, Hu and Scarfone proposed metrics for access control system evaluation or a security system in general~\cite{hu2012:NIST7874}. 
They collect metrics, which can be applicable for not only access control models, but also requirements, system implementation, and extended application such as verification and testing. 
For example, ``Ease of Privilege Assignments'' metric measures the number of steps for assigning, changing, or deleting a privilege for users and roles when a user or administrator manages RBAC. 
This metric can be applicable for not only the RBAC Reference Model but also RBAC extension models.
We propose the metrics introduced by Hu and Scarfone should be applied to the extension models presented in this paper as well as any future models.
A developer could attempt to apply these metrics to an RBAC extension model not already evaluated when attempting to choose a model. 
These metrics focus on comparing feature support of access control models.

@@

To adopt a model in practice, software developers would implement the model in real system environments based on intended use of the models. 
As shown by RQ6, all primary sources were found to provide an abstract formal representations of their extension model and the model's operation.
The presence of an abstract formal representation of the model stands as a starting place for developers to configure intended use (i.e. which subjects can access which resources), and define an access control policy (i.e. which specifies high-level rules such as which subjects can access to which resources).

>>>>

The research questions we identified, presented results for and analyzed provide a view of the RBAC extension model landscape. 
By looking across all the research questions we can arrive at cross-cutting concerns and identify areas that may benefit from future work.
We discuss motivations behind the RBAC extension models and guidelines for choosing an RBAC extension model.

\subsection{Motivations behind the RBAC extension models}

While the RBAC Reference Model is considered fundamental in any RBAC systems, the RBAC Reference Model has limitations on providing features such as dealing with context for emerging applications such as healthcare and mobile devices. For example, in cases where the RBAC extension model does not support the notion of specific constraints, the RBAC extension model incorporates additional entities with regards to context/constraints and their relations with existing entities of the RBAC Reference Model. The RBAC extension model provides general agreement on the definition for specific features. The RBAC extension model helps simplify theoretical modeling and practical implementation of features in the model.

During the paper reading phase, we identify motivations of the RBAC extension models by category. We identify motivations by looking for issues regarding why the RBAC reference model was inadequate and how the authors addressed the issues. We describe motivations behind the RBAC extension models by category as follows:

\begin{itemize}

  \item \textbf{Context}: The RBAC Reference Model does not support the notion of context constraints related to changes in environments. Therefore, RBAC belongs to the static access control model, which may not capture changes in environments.

  \item \textbf{Constraint}: The RBAC Reference Model has limitations on its features such as delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.

  \item \textbf{Organizational}: The RBAC extension model does not handle RBAC administrative tasks efficiently across multiple organizations. The model needs to reduce the administrative complexity of RBAC across multiple organizations.

  \item \textbf{Privacy}: The RBAC extension model does not support the notion of privacy. For example, the model lacks components, constraints, and obligations to handle privacy in RBAC.

  \item \textbf{Task}: The RBAC extension model does not support the notion of a task, team, purpose, and organizational roles, which help specify a business activity in enterprises.

  \item \textbf{Spatio-Temporal}: The RBAC extension model does not support spatial (location-based) and temporal (time-based) constraints, which specify role-assignment, role-activation and permissions based on location and time.

  \item \textbf{Spatial}: The RBAC extension model does not support the notion of spatial (location-based)  constraints, which specify role-assignment and permissions based on location.

  \item \textbf{Temporal}: The RBAC extension model does not support the notion of temporal (time-based)  constraints, which specify role-assignments and permissions based on time.

\end{itemize}

We found that motivations for RBAC extension models vary across category. In general, motivations are rooted in adding new entities or relationships to allow authorization flexibility in meeting the demands of emerging requirements.

@@

To adopt a model in practice, software developers would implement the model in real system environments based on intended use of the models. 
As shown by RQ5, all primary sources were found to provide an abstract formal representations of their extension model and the model's operation.
The presence of an abstract formal representation of the model stands as a starting place for developers to configure intended use (i.e. which subjects can access which resources), and define an access control policy (i.e. which specifies high-level rules such as which subjects can access to which resources).

### 2014-06-17 paper/sections/intro.tex
Revised version
 - Address comments in background, RQ2-RQ5, and discussion section.
 - In addition, I sort paper list in an increasing order (e.g., [4][5][10] instead of [10][4][5]) in the table.

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item RQ1: How can RBAC extension models be classified?
\item RQ2: What are motivations behind the RBAC extension models?
\item RQ3: How are RBAC extension models implemented by their authors?
\item RQ4: How are RBAC extension models evaluated by their authors?
\item RQ5: What domains have been targeted by RBAC extension models?
\item RQ6: What commonalities exist across RBAC extension models?
\end{itemize}

>>>>

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item RQ1: How can RBAC extension models be classified?
%\item RQ2: What are motivations behind the RBAC extension models?
\item RQ2: How are RBAC extension models implemented by their authors?
\item RQ3: How are RBAC extension models evaluated by their authors?
\item RQ4: What domains have been targeted by RBAC extension models?
\item RQ5: What commonalities exist across RBAC extension models?
\end{itemize}
